---
title: "Comparison of three machines"
author: "Quentin D. Read"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Abstract

Here I compare the different machines for measuring total weight. First, I compare within-sample variability for the three machines. We find that the GAC machine has less within-sample variability than the Perten machine, although it suffers from a number of highly variable readings that the other machines do not show. Next I consider differences between machines. The agreement with the standard (Volumetric) machine is significantly better for Perten than it is for GAC. Although both GAC and Perten are biased to give systematically higher total weight readings than Volumetric, this systematic bias is significantly greater for GAC. Finally, I provide the calibration correction equations for GAC and Perten, using Volumetric as a standard, generated using reduced major axis (RMA) regression.

# Setup

Load necessary packages and data. Only the first eight columns are needed.

```{r}
library(data.table)
library(ggplot2)
library(lmodel2)
library(emmeans)
library(multcomp)
library(gt)

theme_set(theme_bw() +
            theme(panel.grid = element_blank(),
                  strip.background = element_blank()))

dat <- fread('Three Machine Anova Individual Measurements.csv', na.strings = '.')[, 1:8]
```

I noticed that trial U7, plot 212, year 2019 has duplicate measurements for some of the reps. I just removed that one to avoid any potential error.

```{r}
dat <- dat[!(Trial %in% 'U7' & Plot %in% 212 & Year %in% 2019)]
```

# Within-machine and between machine variability

To look at within-machine variability for each sample, we will take the average of the absolute value of pairwise differences between measurements on the same sample from the same machine. This is called [Gini's mean difference](https://en.wikipedia.org/wiki/Mean_absolute_difference) (GMD) also known as mean absolute difference.

```{r}
gmd1 <- function(x) mean(abs(dist(x)), na.rm = TRUE) # One sample
gmd2 <- function(x, y) mean(abs(outer(x, y, `-`)), na.rm = TRUE) # Two samples

within_GMD <- dat[, .(GMD = gmd1(TW)), by = .(Trial, Machine, Year, Plot)]
```

For between-machine variability for each sample, we will take the average absolute value of pairwise differences between measurements on the same sample from two different machines. This is Gini's mean difference for two samples. We need to make a wide dataset where each machine has its own column, then get the mean absolute difference for each pair of columns (machines).

```{r}
dat_wide <- dcast(dat, Trial + Year + Plot + Rep ~ Machine, value.var = 'TW') 
between_GMD <- dat_wide[, .(GMD_VG = gmd2(Vol, GAC), GMD_VP = gmd2(Vol, Perten), GMD_GP = gmd2(Vol, Perten)), by = .(Trial, Year, Plot)]
```

## Results: within-machine variability

Notice this boxplot is on a log scale because of the multiple orders of magnitude (actually log+1 because of all the zeros in the data). We see that the Volumetric machine has a slightly higher median value of within-machine variability, but the GAC machine has several extremely high values of variability (> 3) for individual samples which represent a few cases where it gave very bad readings. The Perten machine has a few highly variable data poins but overall performs better.

```{r}
ggplot(within_GMD, aes(y = GMD, x = Machine, fill = Machine)) +
  geom_boxplot(outlier.color = 'gray50', width = 0.6, size = 1) +
  scale_y_continuous(breaks = c(0.1, 0.5, 1, 5, 10), trans = 'log1p') +
  theme(legend.position = 'none') +
  scale_fill_brewer(palette = 'Dark2')
```

Here is an alternative way of showing the same data, as overlapping density plots. The two measurements from GAC with bias >15 are omitted from the plot. Again we see that the Volumetric machine has more within-sample variability.

```{r}
ggplot(within_GMD, aes(x = GMD, fill = Machine)) +
  geom_density(alpha = 0.7) +
  scale_fill_brewer(palette = 'Dark2') +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 5)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.01))) +
  theme(legend.position = c(.8, .8))
```

A statistical test can be used to compare the mean variability for the machines, again we do a log transformation with a small quantity added before analysis, then back-transform when comparing the means. After adjusting for multiple comparisons, the test shows that Volumetric has higher within-sample variability than Perten which has higher within-sample variability than GAC (despite the few high outliers from GAC). Note in the table that the null value being compared to is 1 because a ratio of 1 means no difference when comparing on the log scale.

```{r}
lm_within_GMD <- lm(log(GMD + 0.1) ~ Machine, data = within_GMD)
emm_within_GMD <- emmeans(lm_within_GMD, ~ Machine, type = 'response')

contr_within_GMD <- contrast(emm_within_GMD, 'pairwise', adjust = 'sidak')

cld_within_GMD <- cld(emm_within_GMD, adjust = 'sidak', Letters = letters)
```

```{r, echo = FALSE, results = 'asis'}
contr_dat <- as.data.table(contr_within_GMD)
contr_dat[, p.value := lapply(Rutilitybelt::pretty_sci_not(p.value), html)]

gt(as.data.table(cld_within_GMD)) %>%
  fmt_number(columns = c(2, 3, 5, 6), decimals = 3) %>%
  cols_label(response = 'within-sample variability', lower.CL = '95% CI lower', upper.CL = '95% CI upper', .group = '') %>%
  tab_options(column_labels.font.weight = 'bold')

gt(contr_dat) %>%
  fmt_number(columns = c(2, 3, 6), decimals = 3) %>%
  cols_label(t.ratio = 't ratio', p.value = 'p-value') %>%
  tab_options(column_labels.font.weight = 'bold') %>%
  cols_align('right', columns = p.value)
```


## Results: among-machine variability

Here we see that the GAC machine disagrees more with the Volumetric machine (overall higher values of Gini's mean difference in the Volumetric vs. GAC comparison). The Perten machine agrees better with the Volumetric machine.

```{r}
between_GMD_long <- melt(between_GMD, id.vars = c('Trial', 'Year', 'Plot'), variable.name = 'Pair', value.name = 'GMD')

ggplot(between_GMD_long, aes(y = GMD, x = Pair, fill = Pair)) +
  geom_boxplot(outlier.color = 'gray50', width = 0.6, size = 1) +
  scale_y_continuous(breaks = c(0.1, 0.5, 1, 5, 10), trans = 'log1p') +
  theme(legend.position = 'none') +
  scale_fill_brewer(palette = 'Set1') +
  scale_x_discrete(labels = c('GAC vs. Vol', 'Perten vs. Vol', 'GAC vs. Perten'))
```

# Bias

If we consider the median measurement from the Volumetric machine to be the standard, what is the bias of each machine? Now we will not take the absolute value anymore because we care if it's positive or negative. We will use the *median* value for each machine, not the mean, because the median will not be as sensitive to outliers.

```{r}
median_meas <- dat[, .(median_TW = median(TW, na.rm = TRUE)), by = .(Trial, Machine, Year, Plot)]
bias <- dcast(median_meas, Trial + Year + Plot ~ Machine)
bias[, GV := GAC - Vol]
bias[, PV := Perten - Vol]

bias_long <- melt(bias, id.vars = c('Trial', 'Year', 'Plot'), measure.vars = c('GV', 'PV'), variable.name = 'Pair', value.name = 'bias')
```

Plot the results. Both GAC and Perten are positively biased relative to the Volumetric standard: the majority of the bias values are positive for both. Most biases are less than 2 but there are a few extremely biased measurements in both directions, especially for the GAC machine.

```{r}
ggplot(bias_long, aes(x = Pair, y = bias, fill = Pair)) +
  geom_boxplot(outlier.color = 'gray50', width = 0.6, size = 1) +
  theme(legend.position = 'none') +
  scale_fill_brewer(palette = 'Set1') +
  scale_x_discrete(labels = c('GAC vs. Vol', 'Perten vs. Vol', 'GAC vs. Perten'))
```

What proportion of the biases are positive and negative for both machines? Exclude missing values from this calculation.

```{r}
bias_sums <- bias_long[, .(positive_bias = sum(bias > 0, na.rm = TRUE), negative_bias = sum(bias < 0, na.rm = TRUE), no_bias = sum(bias == 0, na.rm=TRUE)), by = Pair]
bias_props <- sweep(bias_sums[,-1], 1, rowSums(bias_sums[,-1]), `/`)
```

```{r, echo = FALSE}
gt(cbind(Pair = c('GAC - Volumetric', 'Perten - Volumetric'), bias_props)) %>%
  fmt_number(columns = 2:4, decimals = 3) %>%
  cols_label(positive_bias = 'Positive bias', negative_bias = 'Negative bias', no_bias = 'No bias') %>%
  tab_options(column_labels.font.weight = 'bold')
```


Are the biases relative to the Volumetric machine significantly different for GAC versus Perten? We can answer this with a paired t-test. The observed mean is 0.360 (positive value indicates a greater bias for GAC) with a 95% confidence interval of [0.247, 0.474] indicating that GAC has a larger positive bias than Perten. Therefore although GAC and Perten both are positively biased the majority of the time, the bias is bigger for GAC.

```{r}
t.test(Pair(GV, PV) ~ 1, alternative = 'two.sided', var.equal = FALSE, conf.level = 0.95, data = bias)
```


# Calibration equations

We will use median values from each sample to get the calibration equation for GAC and Perten, relative to Volumetric in both cases. We cannot use ordinary least squares (OLS) to fit the model because OLS assumes that x was measured without error and is asymmetric (you get different results from regressing y on x and x on y). But in our case both x and y are measured with error (all three machines have some amount of measurement error). Therefore we will use reduced major axis (RMA) regression to get the slope and intercept of the calibration line. RMA assumes that both x and y were measured with error, and gives the same result no matter which variable is selected to be x and which is y. This is more appropriate than OLS because one machine's measurement does not cause or depend on the other, so it is arbitrary which one is x and which one is y. Therefore it makes more sense to use a symmetrical method. We will also show the OLS results as a comparison. Just like we used for the bias calculations above, we will use the median measurement from each machine to do the regression.

```{r}
fit_V_vs_G <- lmodel2(Vol ~ GAC, data = bias, range.x = 'relative', range.y = 'relative')
fit_V_vs_P <- lmodel2(Vol ~ Perten, data = bias, range.x = 'relative', range.y = 'relative')
```

Plot Volumetric versus GAC and versus Perten. The RMA regression line is given in red (bolded) and the OLS regression line is given in blue. For comparison, a y=x or 45-degree slope line is given as a dashed green line. We want to use the red line to correct GAC and Perten to the standard.

The red lines have a shallower slope than the 1:1 lines, meaning that the correction to standard will tend to reduce the value if the GAC or Perten reading is high, and increase it if it is low. This is more of a pronounced difference for GAC than it is for Perten.

```{r}
ggplot(bias, aes(x = Vol, y = GAC)) +
  geom_point() +
  geom_abline(color = 'red', intercept = fit_V_vs_G$regression.results[4, 'Intercept'], slope = fit_V_vs_G$regression.results[4, 'Slope'], size = 1.2) +
  geom_abline(color = 'blue', intercept = fit_V_vs_G$regression.results[1, 'Intercept'], slope = fit_V_vs_G$regression.results[1, 'Slope']) +
  geom_abline(color = 'green', linetype = 'dashed', intercept = 0, slope = 1) +
  theme(aspect.ratio = 1) + 
  ggtitle('Volumetric versus GAC')

ggplot(bias, aes(x = Vol, y = Perten)) +
  geom_point() +
  geom_abline(color = 'red', intercept = fit_V_vs_P$regression.results[4, 'Intercept'], slope = fit_V_vs_P$regression.results[4, 'Slope'], size = 1.2) +
  geom_abline(color = 'blue', intercept = fit_V_vs_P$regression.results[1, 'Intercept'], slope = fit_V_vs_P$regression.results[1, 'Slope']) +
  geom_abline(color = 'green', linetype = 'dashed', intercept = 0, slope = 1) +
  theme(aspect.ratio = 1) + 
  ggtitle('Volumetric versus Perten')
```

Here are the equations, given to 4 significant figures, from the RMA regression that can be used to convert GAC and Perten readings to standard using the equation

$$TW_{corrected} = a + b TW_{observed}$$

where $a$ is the intercept and $b$ is the slope.

```{r, echo = FALSE}
rma_dat <- cbind(Machine = c('GAC', 'Perten'), rbind(fit_V_vs_G$regression.results[4, 2:3], fit_V_vs_P$regression.results[4, 2:3]))

gt(rma_dat) %>%
  fmt_number(columns = 2:3, n_sigfig = 4) %>%
  tab_options(column_labels.font.weight = 'bold')
```

